Testing for measurement invariance with respect to an ordinal variable

### 2012-10-21 ordinal/ordinal.Rnw
Revisions to ordinal paper: o Added an intermediate, tau-equivalent model to application section o Added a reference
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@172 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%% TODO add some more citations to these initial paragraphs
The study of measurement invariance and differential item functioning
(DIF) 
has received considerable attention in the psychometric literature
(see, e.g., \citeNP{Mil11} for a thorough review).  A set of
psychometric scales $X$ are defined to be measurement invariant
with respect to an auxiliary variable $V$ if \cite{Mel89}
\begin{equation}
    \label{eq:midef}
      f(x_i | t_i, v_i, \dots) = f(x_i | t_i, \dots),
\end{equation}
where $T$ is the latent variable that the scales measure, $f$ is the
model's distributional form, the $i$ subscript refers to
individual cases, capital letters signify random variables, and lowercase
letters signify realizations of the variables.  If the above equation
does not hold, then  
a measurement invariance violation is said to exist.  We focus here on
situations where $f()$ is the probability density function of $X$,
and the measurement invariance
violation occurs because the model parameters are
unequal across individuals (and related to $V$).

@@

In studying measurement invariance, we are essentially testing the
assumption that all  
individuals arise from the same 
parameter vector ${\bm \theta}$.
Thus, a hypothesis of invariance can be written as
\begin{equation}
    \label{eq:h0}
    H_0:~ {\bm \theta}_i = {\bm \theta}_0,\quad (i=1,\ldots,n),
\end{equation}
where ${\bm \theta}_i$ reflects the parameter vector for individual
$i$ (and modifications for subsets of ${\bm \theta}$ are immediate).  
The most general alternative hypothesis related to $V$ may then be
written as
\begin{equation}
  \label{eq:h2*}
  H_1^*:~ {\bm \theta}_i = {\bm \theta}_{v_i},
\end{equation}
stating that the parameter vector differs for every unique realization
of $V$.  This alternative is commonly employed when $V$ is
categorical.  In these situations, the likelihood ratio test (LRT) compares
a model where 
parameters are restricted across groups (i.e., across values of $V$)
to a model where parameters are free across groups; the exact
parameter values within each group are completely unrestricted.
However, in situations where $V$ is categorical or
continuous,~\eqref{eq:h2*} includes non-monotonic violations of
measurement invariance.   This allows for instances where, e.g., the
parameter values initially increase with $V$ and then decrease, 
or where just one or two ``middle'' levels of $V$ differ from the rest.
Researchers typically do not expect such a result when testing
measurement invariance w.r.t.\ continuous or ordinal $V$, and 
researchers often cannot interpret such violations.  Monotonic
parameter changes w.r.t.\ $V$ are of much more interest in these
situations, with the simplest type of change given by the alternative
hypothesis
\begin{equation}
  \label{eq:h1*}
  H_2^*:~ {\bm \theta}_i = \left\{ \begin{array}{ll}
    {\bm \theta}^{(A)} & \mbox{if } v_i \le \nu, \\
    {\bm \theta}^{(B)} & \mbox{if } v_i >   \nu,
  \end{array} \right.
\end{equation}
where $\nu$ is a threshold dividing individuals into two groups based on
$V$.  This alternative is implicitly employed in ``median split''
analyses, % TODO cite preacher
where $\nu$ is given as the sample median of $V$.  The threshold $\nu$
is usually unknown, however, so it is generally of interest to
test~\eqref{eq:h1*} across all possible values of $\nu$.  The tests
proposed below generally allow for this.

@@

To test for measurement invariance w.r.t.\ age, each of the three scales
was individually 
factor-analyzed using the items that comprised the scale.
For each model, the authors 
first fit a congeneric model (all parameters free for each level of age),
followed by a tau-equivalent model (factor loadings restricted to be
equal across each level of age) and a parallel model (all parameters restricted
to be equal across levels of age).  Because their sample size was large ($n
\approx 1400$), they could not rely solely on likelihood ratio tests
(i.e., $\chi^2$ difference tests) for model comparison because the
tests were overly sensitive at their sample size.  To supplement these
tests, \citeA{FroFan11} examined a set of alternative fit indices, including
the non-normed fit index, the comparative fit index, and the
incremental fit index. %% TODO cite
The authors generally found support for the tau-equivalent models
through these alternative fit indices; the likelihood ratio test often
resulted in significance even when the alternative indices indicated
good fit.

@@

Plots representing the statistics' fluctuations across age
groups are displayed in Figure~\ref{fig:gacres}.  The left panel
displays the process associated with $\mathit{WDM}_o$, while the right
panel displays the process associated with $\max\mathit{LM}_o$.  It is
observed that both processes are fully above the critical value,
implying the measurement invariance violation.  Additionally, both
processes peak at the 12--13 age group, suggesting that
parameters differ between individuals up to 13 years of age and older
than 13 years. \fixme{Z: Maybe we could additionally fit a tau-equivalent
model for only two age groups: $\le 13$ vs.\ $> 13$. Then, we could
re-run tests and check whether this already fits sufficiently well?
This could help making the split into two subgroups clearer, rather
than always splitting into all possible age groups.}
\readme{ECM: See code starting on line 963 of the Rnw file. Even if we
  split around age 13, the test statistics are all significant.}
\readme{Z: My formulation above was misleading. We started out from
model m2 with 24 parameters where variances are restricted to be equal
across all groups. The LRT compares this to model m2.f with 44 parameters,
where the group-specific variances for all 6 age groups are estimated.
This ``costs'' us $(6 - 1) \cdot 4 = 20$ additional parameters. Instead,
we could fit a model m2.2 with 4 additional parameters -- i.e., 28 in total --
where the variances are just estimated separately for two age groups (split at age 14).
This should improve significantly upon m2 but if we are lucky it is not
significantly worse than m2.f.}

<<gac.2grp>>=
yg <- transform(yg, under14=as.numeric(agegroup)<=2)

m3 <- cfa('f1 =~ gac_1 + gac_2 + gac_3
           f1 ~ 0*1', data=yg, group="under14",
           group.equal=c("loadings","residuals","lv.variances"), 
           meanstructure=TRUE)
gefp3 <- gefp(m3, fit = NULL, scores=estfun.lavaan, 
              order.by = as.numeric(yg$under14), 
              vcov = info_full, sandwich = FALSE, parm=3:6)
if(file.exists("ol2bb3.rda")){
  load("ol2bb3.rda")
} else {
  ol2bb3.fun <- ordL2BB(gefp3)
  save(ol2bb3.fun, file="ol2bb3.rda")
}
omax3 <- sctest(gefp3, functional=ordwmax(gefp3))
ol2bb3 <- sctest(gefp3, functional=ol2bb3.fun)

## Full model
m3.f <- cfa('f1 =~ gac_1 + gac_2 + gac_3
           f1 ~ 0*1', data=yg, group="under14",
           group.equal="loadings", meanstructure=TRUE)
lrt3 <- anova(m3.f, m3)
@

>>>>

The study of measurement invariance and differential item functioning
(DIF) 
has received considerable attention in the psychometric literature
(see, e.g., \citeNP{Mil11} for a thorough review).  A set of
psychometric scales $X$ are defined to be measurement invariant
with respect to an auxiliary variable $V$ if \cite{Mel89}
\begin{equation}
    \label{eq:midef}
      f(x_i | t_i, v_i, \dots) = f(x_i | t_i, \dots),
\end{equation}
where $T$ is the latent variable that the scales measure, $f$ is the
model's distributional form, the $i$ subscript refers to
individual cases, capital letters signify random variables, and lowercase
letters signify realizations of the variables.  If the above equation
does not hold, then  
a measurement invariance violation is said to exist.  We focus here on
situations where $f()$ is the probability density function of $X$,
and the measurement invariance
violation occurs because the model parameters are
unequal across individuals (and related to $V$).

@@

In studying measurement invariance, we are essentially testing the
assumption that all  
individuals arise from the same 
parameter vector ${\bm \theta}$.
Thus, a hypothesis of invariance can be written as
\begin{equation}
    \label{eq:h0}
    H_0:~ {\bm \theta}_i = {\bm \theta}_0,\quad (i=1,\ldots,n),
\end{equation}
where ${\bm \theta}_i$ reflects the parameter vector for individual
$i$ (and modifications for subsets of ${\bm \theta}$ are immediate).  
The most general alternative hypothesis related to $V$ may then be
written as
\begin{equation}
  \label{eq:h2*}
  H_1^*:~ {\bm \theta}_i = {\bm \theta}_{v_i},
\end{equation}
stating that the parameter vector differs for every unique realization
of $V$.  This alternative is commonly employed when $V$ is
categorical.  In these situations, the likelihood ratio test (LRT) compares
a model where 
parameters are restricted across groups (i.e., across values of $V$)
to a model where parameters are free across groups; the exact
parameter values within each group are completely unrestricted.
However, in situations where $V$ is categorical or
continuous,~\eqref{eq:h2*} includes non-monotonic violations of
measurement invariance.   This allows for instances where, e.g., the
parameter values initially increase with $V$ and then decrease, 
or where just one or two ``middle'' levels of $V$ differ from the rest.
Researchers typically do not expect such a result when testing
measurement invariance w.r.t.\ continuous or ordinal $V$, and 
researchers often cannot interpret such violations.  Monotonic
parameter changes w.r.t.\ $V$ are of much more interest in these
situations, with the simplest type of change given by the alternative
hypothesis
\begin{equation}
  \label{eq:h1*}
  H_2^*:~ {\bm \theta}_i = \left\{ \begin{array}{ll}
    {\bm \theta}^{(A)} & \mbox{if } v_i \le \nu, \\
    {\bm \theta}^{(B)} & \mbox{if } v_i >   \nu,
  \end{array} \right.
\end{equation}
where $\nu$ is a threshold dividing individuals into two groups based on
$V$.  This alternative is implicitly employed in ``median split''
analyses, 
where $\nu$ is given as the sample median of $V$.  The threshold $\nu$
is usually unknown, however, so it is generally of interest to
test~\eqref{eq:h1*} across all possible values of $\nu$.  The tests
proposed below generally allow for this.

@@

To test for measurement invariance w.r.t.\ age, each of the three scales
was individually 
factor-analyzed using the items that comprised the scale.
For each model, the authors 
first fit a congeneric model (all parameters free for each level of age),
followed by a tau-equivalent model (factor loadings restricted to be
equal across each level of age) and a parallel model (all parameters restricted
to be equal across levels of age).  Because their sample size was large ($n
\approx 1400$), they could not rely solely on likelihood ratio tests
(i.e., $\chi^2$ difference tests) for model comparison because the
tests were overly sensitive at their sample size.  To supplement these
tests, \citeA{FroFan11} examined a set of alternative fit indices, including
the non-normed fit index, the comparative fit index, and the
incremental fit index (e.g., \citeNP{BroCud93}).
The authors generally found support for the tau-equivalent models
through these alternative fit indices; the likelihood ratio test often
resulted in significance even when the alternative indices indicated
good fit.

@@

Plots representing the statistics' fluctuations across age
groups are displayed in Figure~\ref{fig:gacres}.  The left panel
displays the process associated with $\mathit{WDM}_o$, while the right
panel displays the process associated with $\max\mathit{LM}_o$.  It is
observed that both processes are fully above the critical value,
implying the measurement invariance violation.  Additionally, both
processes peak at the 12--13 age group, suggesting that
parameters differ between individuals up to 13 years of age and
individuals older 
than 13 years of age.

<<gac.mod>>=
m2.2 <- cfa('f1 =~ gac_1 + gac_2 + gac_3
             f1 ~ 0*1
             gac_1 ~~ c("v1.1","v1.1","v1.2","v1.2","v1.2","v1.2")*gac_1
             gac_2 ~~ c("v2.1","v2.1","v2.2","v2.2","v2.2","v2.2")*gac_2
             gac_3 ~~ c("v3.1","v3.1","v3.2","v3.2","v3.2","v3.2")*gac_3
             f1 ~~ c("v4.1","v4.1","v4.2","v4.2","v4.2","v4.2")*f1', 
             data=yg, group="agegroup",
             group.equal=c("loadings","residuals","lv.variances"), 
             meanstructure=TRUE)

gefp2.2 <- gefp(m2.2, fit = NULL, scores=estfun.lavaan, 
                order.by = as.numeric(yg$agegroup), 
                vcov = info_full, sandwich = FALSE, parm=c(3:6,13:16))

if(file.exists("ol2bb22.rda")){
  load("ol2bb22.rda")
} else {
  ol2bb22.fun <- ordL2BB(gefp2.2)
  save(ol2bb22.fun, file="ol2bb22.rda")
}

omax2.2 <- sctest(gefp2.2, functional=ordwmax(gefp2.2))
ol2bb2.2 <- sctest(gefp2.2, functional=ol2bb22.fun)
lrt2.2par <- anova(m2.2, m2)
lrt2.2te <- anova(m2.f, m2.2)
@

The finding that variance parameters differ between individuals up to 13 years and individuals over 13 years is reinforced by comparing the
tau-equivalent and parallel models to an intermediate model.  This intermediate model is tau-equivalent in nature, but there exist only two groups: individuals up to 13 years
and individuals older than 13 years.  A likelihood ratio test implies that
this intermediate model fits as well as the original tau-equivalent model~($\chi^2_{16} = \Sexpr{round(lrt2.2te[[5]][2], digits = 2)}$, $p=\Sexpr{round(lrt2.2te[[7]][2], digits=2)}$), with 16 fewer parameters.  The intermediate model also fits 
better than the parallel model, as judged by a second likelihood ratio test~($\chi^2_{4} = \Sexpr{round(lrt2.2par[[5]][2], digits = 2)}, p<.01$).
Finally, using the proposed ordinal test statistics with the intermediate model, we no 
longer observe instability in the variance parameters
($\mathit{WDM}_o=\Sexpr{round(omax2.2$statistic, digits = 2)}, p=\Sexpr{round(omax2.2$p.value, digits = 2)}$; 
$\max\mathit{LM}_o=\Sexpr{round(ol2bb2.2$statistic, digits = 2)}, p=\Sexpr{round(ol2bb2.2$p.value, digits = 2)}$).
